Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses

### 2021-12-30 mk3/manuscript.rmd
added revision 3

---
title: "Erroneous use of Venn diagrams leads to artifacts in transcriptomic analyses"
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

Venn diagrams are frequently used to illustrate the results of
differential expression analysis and further
downstream functional analysis. Here we show that both, the use
of Venn diagrams to find genes which are thought to be specific for a
certain comparison, as well as gene set enrichment analysis applied to such
subsets is a fallacy. Since genes which show a significant change in one
condition but not in another are likely to be false negatives in the latter
case, these genes considered to be specific for one condition are likely
characteristic for both of the compared groups, and enriched in functions
related to the research hypothesis. Thus, the combination of incorrect
statistical reasoning and gene set enrichment analysis may result in
particularly misleading artifacts.

@@

```{r libraries,cache=FALSE}
library(tidyverse)
library(limma)
library(GEOquery)
library(DESeq2)
## install bioshmods from github.com/bihealth/bioshmods
library(bioshmods)
library(tmod)
library(cowplot)
library(stringr)
library(ggvenn)
library(ggplot2)
library(ggpval)
library(eulerr)
library(Biobase)
library(org.Hs.eg.db)
theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
source("functions.R")
```

@@

Venn diagrams (VDs) are commonly used to visualize high throughput data
such as transcriptomic profiles. Frequently, a Venn diagram serves the
purpose of comparing the results from two experiments or two comparisons.
For example, a VD may illustrate up- or down-regulated genes in two strains
of mice upon infection, showing which genes are regulated in both strains,
and which are regulated only in one of them.  The VD serves
as a basis for the statement that the regulation of certain genes is
"specific" for one strain or another. We argue that not only is this
inference incorrect, but it may also lead to misleading – although
appealing – artifacts when combined with downstream analyses.

@@

Consider two group of patients, G1 and G2. Each group contains 
`r n_stratum * 4`
individuals. In both groups, there is an equal number of healthy
individuals (labeled "Ctrl" on figures below) 
or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
the differences between G1 and G2 in the response to infection. For
example, we ask which genes or pathways are specifically upregulated by SC2
infection in G1 as compared to G2, and vice versa. In the following, we
used the data set GSE156063 [@mick2020upper] in two approaches (an
incorrect and the correct one) which arrive at opposite conclusions.

@@

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis (GSEA) – a hypergeometric test – on the DEGs in each
group. Gene sets for GSEA were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding "specifically" regulated pathways.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

@@

plot_grid(part1, part2, rel_widths=c(2, 2), labels=c("", "E"))
```
Consequently, it is not possible to calculate gene set enrichment for the
interaction using a
hypergeometric test, as there are no DEGs for the interaction contrast.
Gene set enrichment using a second generation algorithm (CERNO), relying on the
ordering of genes according to their raw p-values from the interaction
contrast rather than selecting a
set of DEGs [@zyla2019gene], does not show any significant enrichment.

@@

```{r lit_surves}
links <- system2("grep", args=c("https", "literature_survey.md"), stdout=TRUE)
n_tot <- length(links)
n_vd_only <- length(grep("vd only", links))
```

In an informal survey of articles from the journal "Scientific reports"
(chosen for its broad scope, large number of published articles and open
access), of the 238 articles from 2021 which used the term "venn diagram" and
"differential expression", at least 
`r n_tot`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r n_vd_only`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs.

@@

As an alternative to Venn diagrams two approaches can be considered.
Firstly, as shown above, a test for interaction can reveal genes for which
the impact of treatment significantly differ between the groups. As a mean
of visualization, we recommend plotting the log~2~ fold changes in one
comparison versus the log~2~ fold changes in the other comparison. In
addition, a concordance / discordance analysis may provide additional
insight in genes which are regulated either in the same direction or in the
opposite directions in both groups [@domaszewska2017concordant].

While the use of Venn diagrams in combination with the incorrect gene set
enrichment analysis is common, it is not the only way of achieving
artifactual and incorrect conclusions by means of comparing significant
with non-significant results in transcriptomics and other high-throughput
applications. Our literature survey showed that at least two other ways are
common. Firstly, the direct comparison of gene set enrichment
results: that is, drawing conclusions from the fact that a gene set
enrichment result was significant in one comparison only.
Second, while Venn diagrams are often used to illustrate the numbers of
"specific" DEGs and so present a mean to find examples of this fallacy in
scientific literature, researches often test for enrichment these
"specific" genes without using the phrase "Venn diagram" or even clearly
stating how the lists of "specific" genes were derived. In all these cases,
the analysis boils down to comparing results significant in one, but not in
another comparison.

>>>>

---
title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
output:
  html_document
outputoff:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

Venn diagrams are frequently used to illustrate the results of
differential expression analysis and further
downstream functional analysis. Here we show that both, the use
of Venn diagrams to find genes which are thought to be specific for a
certain comparison, as well as gene set enrichment analysis applied to such
subsets is a fallacy.  A statistically correct approach involves testing for interaction.
Moreover, since genes which show a significant change in one condition but
not in another are likely to be false negatives in the latter case, these
genes considered to be specific for one condition are likely characteristic
for both of the compared groups, and enriched in functions related to the
research hypothesis. Thus, the combination of incorrect statistical
analysis and gene set enrichment analysis may result in particularly
misleading artifacts.

@@

```{r libraries,cache=FALSE}
library(tidyverse)
library(limma)
library(GEOquery)
library(DESeq2)
## install bioshmods from github.com/bihealth/bioshmods
library(bioshmods)
library(tmod)
library(cowplot)
library(stringr)
library(ggvenn)
library(ggplot2)
library(ggpval)
library(eulerr)
library(Biobase)
library(pander)
library(org.Hs.eg.db)
library(pwr)
theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
source("functions.R")
```

@@

Venn diagrams (VDs) are commonly used to visualize high throughput data
such as transcriptomic profiles. Frequently, a Venn diagram serves the
purpose of comparing the results from two experiments or two comparisons.
For example, a VD may illustrate up- or down-regulated genes in two strains
of mice upon infection, showing which genes are regulated in both strains,
and which are regulated only in one of them.  The VD serves as a basis for
the statement that the regulation of certain genes is "specific" for one
strain or another. A gene is, in this context, considered "specific" or
"unique" for a given comparison when it shows a statistically significant
difference in that comparison only.  We argue that not only is this
inference incorrect, but it may also lead to misleading – although
appealing – artifacts when combined with downstream analyses.

@@

Consider two group of patients, G1 and G2 (Tab. 1). Each group contains 
`r n_stratum * 4`
individuals. In both groups, there is an equal number of healthy
individuals (labeled "Ctrl" on figures below) 
or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
the differences between G1 and G2 in the response to infection. For
example, we ask which genes or pathways are specifically upregulated by SC2
infection in G1 as compared to G2, and vice versa. In the following, we
used the data set GSE156063 [@mick2020upper] in two approaches (an
incorrect and the correct one) which arrive at opposite conclusions.

**Table 1** Overall design in the case study: transcriptomic changes due to
Sars-Cov-2 infection. The table shows number of patients in each
combination of study group / disease status.

```{r results="markdown"}
tab <- as.data.frame(matrix(2 * n_stratum, ncol=2, nrow=2))
tab <- data.frame(c("", "Disease status", ""), c("", "Sars-Cov-2 infection", "Another infection"), 
  c("Group 1 (G1)", tab[[1]]), c("Group 2 (G2)", tab[[2]]))

colnames(tab) <- c("", "", "Study group", "")
pander(tab)
```

@@

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis – a hypergeometric test – on the DEGs in each
group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding "specifically" regulated pathways.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

@@

plot_grid(part1, part2, rel_widths=c(2, 2), labels=c("", "E"))
```

Consequently, it is not possible to calculate gene set enrichment for the
interaction using a
hypergeometric test, as there are no DEGs for the interaction contrast.
Gene set enrichment using a second generation algorithm (CERNO), relying on the
ordering of genes according to their raw p-values from the interaction
contrast rather than selecting a
set of DEGs [@zyla2019gene], does not show any significant enrichment.

@@

```{r lit_surveys}
get_keywords <- function(file, pattern="^([a-zA-Z]+): *(.*)") {
  ll <- readLines(file)
  ll <- ll[ grepl(pattern, ll) & !grepl("^https://", ll) ]

  ret <- gsub(pattern, "\\2", ll)
  names(ret) <- gsub(pattern, "\\1", ll)
  ret <- as.list(ret)
  ret$file <- file
  ret$incorrect_count <- get_incorrect_counts(file)
  ret$incorrect_count_vdonly <- get_incorrect_counts(file, vdonly=TRUE)
  ret
}

get_incorrect_counts <- function(file, vdonly=FALSE) {

  ll <- readLines(file)

  pat <- "^https://"
  if(vdonly) {
    sum(grepl(pat, ll) & grepl("vd only", ll))
  } else {
    sum(grepl(pat, ll))
  }

}

files <- list.files(pattern="literature_survey.*.md")

finfo <- map(files, get_keywords) 
names(finfo) <- map_chr(finfo, ~ .x[["journal"]])
```

**Tab. 2** Results of the informal literature survey. 
We searched for papers using Google Scholar and the keywords "venn diagram"
and "differential expression".
*Journal*, journal title; *Years*, publication dates; *Total*, total number
of papers found using the search phrase; *Analysed*, total number of papers
which have been analysed for correctness; *Incorrect*, total number
(percentage) of
papers in which the Venn diagram was combined with comparing significance
to non-significance; *Enrichment*, total numebr (percentage) of papers
which combined the Venn diagram with a gene set enrichment analysis.

```{r results="markdown"}
tab <- map_dfr(finfo, ~ 
               data.frame(Journal=.x$journal,
                          Years=.x$years,
                          Total=.x$total,
                          Analysed=.x$analysed,
                          Incorrect=.x$incorrect_count,
                          Enrichment=.x$incorrect_count - .x$incorrect_count_vdonly))  %>%
  mutate(Analysed=as.numeric(Analysed)) %>%
  mutate(Total=as.numeric(Total))

tab <- rbind(tab, 
             data.frame(Journal="**Total:**", Years="", Total=sum(tab$Total),
                        Analysed=sum(tab$Analysed),
                        Incorrect=sum(tab$Incorrect),
                        Enrichment=sum(tab$Enrichment)))
tab2 <- tab %>%
  mutate(Incorrect=sprintf("%d (%.0f%%)", Incorrect, 100 * Incorrect / Analysed)) %>%
  mutate(Enrichment=sprintf("%d (%.0f%%)", Enrichment, 100 * Enrichment / Analysed))

pander(tab2)
```

In an informal survey of articles from three journals (Tab. 2) we found
that of the 
`r tab2$Analysed[nrow(tab2)]` analysed articles
 which used the term "venn diagram" and
"differential expression", at least 
`r tab2$Incorrect[nrow(tab2)]`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r tab$Enrichment[nrow(tab2)]`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs.

@@

As an alternative to Venn diagrams and the downstream gene set enrichment
analysis, two approaches can be considered.

Statistical analysis, as shown above, involves a test for interaction which
can reveal genes for which the impact of treatment significantly differ
between the groups. The results can then be plugged into a gene set
enrichment analysis the usual way.  Unfortunately, this has two major
drawbacks. Firstly, effect sizes (log~2~ fold changes) of the interaction
term are harder to interpret than log~2~ fold changes in a direct, group vs
group comparison.  The effect size in an interaction is negative if the
log~2~ fold change in the first comparison is larger than the log~2~ fold
change in the second comparison; this is, however, irrespective of whether
the differences in the individual comparisons are negative or positive,
which makes it harder to separate the differences in genes up-regulated in
one or both groups.

The second problem may arise if the changes are similar in both
comparisons, but of larger magnitude in one of them. For example, in a time
series context, the changes may be more pronounced at a later time point.
In this case, the analysis will show that the processes enriched for the
interaction term are the same as those enriched in each of the comparisons
individually.  While the results of the gene set enrichment analysis in
this context are correct, the result may not be what the researchers
intended – processes which qualitatively (rather than quantitatively)
differ between the comparisons.

An alternative approach, discordance/concordance analysis, has been proposed by @domaszewska2017concordant,
aiming at identifying processes which qualitatively differ between the two
comparisons. Here, a heuristic score ("disco score") has been defined which depends on the
effect sizes and p-values in both comparisons. The sign of the score
depends on whether the effects have the same sign (concordant; genes
up-regulated in both comparisons or down-regulated in both comparisons) or
opposite signs (discordant; genes up-regulated in one, but down-regulated
in the other comparison, and genes down-regulated in one, but up-regulated
in the other comparison). While the score does not allow the calculation of
a p-value and does not present an alternative to an analysis of
interaction, it can facilitate both visualization and further analysis
using a gene set enrichment algorithm.

Visualization of interaction for individual genes is straightforward (see
for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
overview of the whole analysis summarising thousands of results for the
analysed genes. As an alternative of such an overview, we suggest plotting
the log~2 fold changes in one comparison against log~2~ fold changes in the
second comparison. This allows an intuitive assessment of the differences
between the two comparisons, in especially with combination of color coding
the genes either are significant in the interaction or using the disco
score (see Fig. 2, *E* for an example).

While the use of VDs in combination with the incorrect gene set
enrichment analysis is common, it is not the only way of achieving
artifactual and incorrect conclusions by means of comparing significant
with non-significant results in transcriptomics and other high-throughput
applications. VDs appear to be symptomatic of the assumption that if a gene
is significantly regulated in one comparison, but does not show a
significant difference in another one, then this gene is "specific" to this
first comparison. However, VDs may also lead one to this reasoning; for
example, the VD on Fig. 1A practically begs the question: what do these 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
genes, which are significant in G1 only, have in common?

Our literature survey showed that even in cases where a VD is absent from a
paper, at least two other ways are not uncommon. Firstly, the direct
comparison of gene set enrichment results: that is, drawing conclusions
from the fact that a gene set enrichment result was significant in one
comparison only.  Second, while Venn diagrams are often used to illustrate
the numbers of "specific" DEGs and so present a mean to find examples of
this fallacy in scientific literature, researches often test for enrichment
these "specific" genes without using the phrase "Venn diagram" or even
clearly stating how the lists of "specific" genes were derived. In all
these cases, the analysis boils down to comparing results significant in
one, but not in another comparison.
